Two subtle problems with over-representation analysis

### 2024-06-13 manuscript/manuscript.Rmd
ms

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
where the results appear nearly instantaneously.
On the other hand, FCS typically involves installing specific softwares and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

Despite the popularity of ORA, there are concerns that this technique is being misused.
Due to technological and biological reasons, not all genes are detected in transcriptomic
studies, meaning that some genes will more readily appear as differentially expressed.
To address this bias, a background list of detected genes is required for comparison to the
list of selected genes (foreground list) [@Tilford2009-vx].
Failure to use a background gene list leads to dramatic changes in enrichment results,
rendering them unreliable [@Timmons2015-ki].
Unfortunately, use of an appropriate background list is reported in only a small fraction
(~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
Correction of p-values for multiple testing is also crucial in controlling the false positive rate
as enrichment analysis typically involves throusands of parallel tests.
But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
These issues have resulted in the development of best practices for end users and
stronger reporting standards [@Zhao2023-mx].

>>>>

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
with results appearing nearly instantly.
On the other hand, FCS typically involves installing specific softwares and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

Despite the popularity of ORA, there are concerns that this technique is being misused.
Due to technological and biological reasons, not all genes are detected in transcriptomic
studies, meaning that some genes will more readily appear as differentially expressed.
To address this bias, a background list of detected genes is required for comparison to the
list of selected genes (foreground list) [@Tilford2009-vx].
Failure to use a background gene list leads to dramatic changes in enrichment results,
rendering them unreliable [@Timmons2015-ki;@Wijesooriya2022-fw].
Unfortunately, use of an appropriate background list is reported in only a small fraction
(~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
Correction of p-values for multiple testing is also crucial in controlling the false positive
rate as enrichment analysis typically involves throusands of parallel tests [@Tilford2009-vx].
But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
These issues have resulted in the development of best practices for end users and
stronger reporting standards [@Zhao2023-mx].
